---
title: "Reproducibility in corpus-based computational analyses of learner speech - Part 2"
author: "Martin Schweinberger"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: word_document
---

# Introduction

This R Notebook shows how to access Praat from R and how to extract formant values from wav files and their corresponding Praat TextGrids. The tutorials is based on [Phonetics Tools in R](https://marissabarlaz.github.io/portfolio/phoneticstools/] by Marissa Barlaz).

## Preparation

install packages

```{r install, eval=F, message=F, warning=F}
# install
install.packages("tidyverse")
install.packages("here")
install.packages("viridis")
install.packages("phonR")
install.packages("vowels")
install.packages("ggridges")
install.packages("sm")
```

load packages

```{r load, message=F, warning=F}
library(tidyverse)
library(here)
library(viridis)
library(phonR)
library(vowels)
library(ggridges)
library(sm)
options("scipen" = 100, "digits" = 12) # suppress math annotation
```


load data


```{r data, message=F, warning=F}
# load .rda data
vdf  <- base::readRDS(file = here::here("data", "vdf.rda")) %>%
  dplyr::mutate(type = stringr::str_replace_all(file, "^[A-Z]{2,}_([A-Z]{3,3})_.*", "\\1"),
                speaker = stringr::str_replace_all(file, ".*_([0-9]{3,3})_.*", "\\1")) %>%
  dplyr::mutate(tgender = stringr::str_replace_all(path, ".*F[3-7]{1,1}_([fm]{1,1})_[a-z]{2,2}$", "\\1"),
                tformants = stringr::str_replace_all(path, ".*(F[3-7]{1,1})_[fm]{1,1}_[a-z]{2,2}$", "\\1"),
                tvariety = stringr::str_remove_all(path, ".*_"))
# inspect
nrow(vdf); head(vdf)
```


# Load Biodata

```{r bio, message=F, warning=F}
bio <- read.delim(here::here("data", "biodata_icnale.txt"), header = T, sep = "\t") %>%
  dplyr::filter(stringr::str_detect(Code, "JPN|ENS")) %>%
  dplyr::mutate(type = stringr::str_replace_all(Code, "^[A-Z]{2,}_([A-Z]{3,3})_.*", "\\1"),
                speaker = stringr::str_replace_all(Code, ".*_([0-9]{3,3}).*", "\\1"),
                Sex = tolower(Sex)) %>%
  dplyr::rename(gender = Sex)
bio[bio=="N/A"]<-NA
# inspect
head(bio)
```


# Join data sets

Join formant and biodata

```{r combine, message=F, warning=F}
fdat <- dplyr::left_join(vdf, bio, by = c("type", "speaker"))  %>%
  # filter out di- and triphthongs
  dplyr::mutate(nv = nchar(stringr::str_remove_all(label, ":"))) %>%
  dplyr::filter(nv == 1) %>%
  dplyr::filter(gender == tgender) %>%
  # remove  ref speakers with l1-var other than genam or gb
  dplyr::mutate(tvariety = dplyr::case_when(Country == "ENS_GBR" ~ "gb",
                                            Country == "ENS_CAN" ~ "us",
                                            Country == "ENS_USA" ~ "us",
                                             TRUE ~ tvariety)) %>%
  dplyr::filter(type == "JPN" | type == "ENS" & Country == "ENS_CAN" | Country == "ENS_GBR" | Country == "ENS_USA")
# inspect
head(fdat)
```






testing

```{r}
table(fdat$label, fdat$tvariety, fdat$type)
```

check proficiency




```{r}
fdat <- fdat %>%
  dplyr::mutate(prof = stringr::str_replace_all(file, ".*_([A-B]{1,1}[0-9]{1,1})_.*", "\\1")) %>%
  mutate(prof = ifelse(prof == file, NA, prof),
         prof = factor(prof, levels = c("A2", "B1", "B2")))
# inspect
summary(fdat$prof)
```



# Reference data

**GenAm**

[Yang, Byunggon. 1996. A comparative study of American English and Korean vowels produced by male and female speakers. *Journal of Phonetics* 24: 245–261.](http://fonetiks.info/bgyang/db/96jp.pdf)

```{r refame, message=F, warning=F}
ame <- read.delim(here::here("data/reference", "GenAm.txt"), sep = "\t", header = T) %>%
  dplyr::mutate(label = dplyr::case_when(item == "had" ~ "{",
                                         item == "heed" ~ "i:",
                                         item == "hid" ~ "I",
                                         item == "head" ~ "E",
                                         item == "hard" ~ "6",
                                         item == "whod" ~ "u:",
                                         item == "hood" ~ "U",
                                         item == "hoard" ~ "O:",
                                         item == "hud" ~ "V",
                                         item == "hod" ~ "Q",
                                         TRUE ~ "other"),
                gender = dplyr::case_when(sex == "female" ~ "f",
                                          TRUE ~ "m")) %>%
  dplyr::filter(label != "other") %>%
  dplyr::group_by(label, gender) %>%
  dplyr::summarise(target_f1 = mean(F1_Hz),
                   target_f2 = mean(F2_Hz),
                   target_f3 = mean(F3_Hz)) %>%
  dplyr::mutate(tvariety = "us")
# inspect data
head(ame)
```

**SSB**

[Deterding, David 2009. The Formants of Monophthong Vowels in Standard Southern British English Pronunciation. *Journal of the International Phonetic Association* 27, 1-2: 47-55. doi:10.1017/S0025100300005417](http://journals.cambridge.org/download.php?file=%2F7775_45AD5CD7849D3B7634A613DE6E61C294_journals__IPA_IPA27_1-2_S0025100300005417a.pdf&cover=Y&code=0639c34593b21cb18a9e7b504bdd5319)




```{r refbre, message=F, warning=F}
bre <- read.delim(here::here("data/reference", "SSB.txt"), sep = "\t", header = T) %>%
  dplyr::mutate(label = dplyr::case_when(item == "had" ~ "{",
                                         item == "heed" ~ "i:",
                                         item == "hid" ~ "I",
                                         item == "head" ~ "e",
                                         item == "hard" ~ "A:",
                                         item == "whod" ~ "u:",
                                         item == "hood" ~ "U",
                                         item == "hoard" ~ "O:",
                                         item == "hud" ~ "V",
                                         item == "hod" ~ "Q",
                                         TRUE ~ "other"),
                gender = dplyr::case_when(sex == "female" ~ "f",
                                          TRUE ~ "m")) %>%
  dplyr::filter(label != "other") %>%
  dplyr::group_by(label, gender) %>%
  dplyr::summarise(target_f1 = mean(F1_Hz),
                   target_f2 = mean(F2_Hz),
                   target_f3 = mean(F3_Hz)) %>%
  dplyr::mutate(tvariety = "gb")
# inspect data
head(bre)
```

Combine reference and obs data

```{r cref, message=F, warning=F}
ref <- rbind(ame, bre)
fdat <- dplyr::left_join(fdat, ref) %>%
  dplyr::filter(!is.na(target_f1)) %>%
  dplyr::mutate(edist = sqrt((F1-target_f1)^2 + (F2-target_f2)^2))
# inspect
head(fdat)
```


Determine best fit (f3:f7)


```{r fit, message=F, warning=F}
fdat <- fdat %>%
  dplyr::arrange(file, tmin, tvariety, tformants) %>%
  dplyr::group_by(file, tmin) %>% 
  dplyr::slice(which.min(edist))
# inspect
head(fdat)
```



remove vowels other than I and i: or U and u:

```{r}
# reduce words
fdat <- fdat %>%
  dplyr::filter(label == "U" | label == "u:" | label == "I" | label == "i:") %>%
  dplyr::filter(nchar(word) < 9,
                word != "the") %>%
  # bark normalization
  dplyr::mutate(barkF1 = normBark(F1),
                barkF2 = normBark(F2)) %>%
  # lobanov normalization
  dplyr::ungroup() %>%
  dplyr::mutate(fspeaker = paste0(type, speaker, collaspe = "")) %>%
  dplyr::group_by(type, gender, tvariety) %>%
  dplyr::mutate(lobF1 = scale(F1),
                lobF2 = scale(F2)) %>%
  # remove superfluous columns
  dplyr::select(-path, -PTJ1..wds., -PTJ2..wds., -SMK1..wds., -SMK2..wds., -Self.Ev, 
                -Grade.Degree, -Acad..Genre, -VST, -CEFR, -INTM, -INSM, -INTM.INSM, 
                -INTM.INSM.1, -Primary, -Secondary, -College, -Inschool, -Outschool,
                -Reading, -Writing, -NS, -Presentation, -EssayW, -nv)
# inspect
head(fdat)
```



# (Semi-)Manual cross-evaluation

check for inconsistencies

```{r check, message=F, warning=F}
table(fdat$label, fdat$word)
```

exclude inconsistencies

```{r exclude, message=F, warning=F}
fdat <- fdat %>%
  dplyr::mutate(word = stringr::str_squish(word)) %>%
  dplyr::mutate(remove = dplyr::case_when(word == "--we"  ~ "remove",
                                          word == "ba"  ~ "remove",
                                          word == "bit" & label == "6" ~ "remove",
                                          word == "books" & label == "I" ~ "remove",
                                          word == "but" & label == "I" ~ "remove",
                                          word == "bui" & label == "I" ~ "remove",
                                          word == "de" & label == "i:" ~ "remove",
                                          word == "do" & label == "6" ~ "remove",
                                          word == "do" & label == "I" ~ "remove",
                                          word == "dust" & label == "V" ~ "remove",
                                          word == "get" & label == "I" ~ "remove",
                                          word == "gets" & label == "I" ~ "remove",
                                          word == "good" & label == "I" ~ "remove",
                                          word == "hea" ~ "remove",
                                          word == "peo" ~ "remove",
                                          word == "speak" & label == "I" ~ "remove",
                                          word == "sets" & label == "I" ~ "remove",
                                          word == "shops" & label == "I" ~ "remove",
                                          word == "so" & label == "U" ~ "remove",
                                          word == "that" & label == "I" ~ "remove",
                                          word == "that's" & label == "I" ~ "remove",
                                          word == "they're" ~ "remove",
                                          word == "they" ~ "remove",
                                          word == "that\u0081's" & label == "I" ~ "remove",
                                          word == "this" & label == "O:" ~ "remove",
                                          word == "to" ~ "remove",
                                          word == "too" & label == "I" ~ "remove",
                                          word == "too" & label == "6" ~ "remove",
                                          word == "two" & label != "u:" ~ "remove",
                                          word == "ways" & label == "I" ~ "remove",
                                          word == "we're" & label == "6" ~ "remove",
                                          word == "we're" & label == "6" ~ "remove",
                                          stringr::str_detect(word, "\\W") ~ "remove",
                                          TRUE ~ "keep")) %>%
  dplyr::filter(remove != "remove")
# inspect
table(fdat$label, fdat$word)
```



## Check durations

```{r}
fdat <- fdat %>%
  # duration
  dplyr::mutate(duration = tmax - tmin)
# inspect
head(fdat)
```

plot durations


```{r}
fdat %>%
  ggplot(aes(x = vowel, y = duration, fill = type)) +
  geom_boxplot()
```







```{r}
max_durs <- fdat %>%
  dplyr::group_by(type, tvariety, vowel) %>%
  dplyr::summarise(du_min = min(duration),
                   du_max = max(duration),
                   du_ave = mean(duration),
                   du_med = median(duration),
                   du_sd = sd(duration),
                   du_2sd = sd(duration)*1.645) %>%
  dplyr::mutate(cutoff_duration = du_ave + du_2sd) %>%
  dplyr::select(type, tvariety, vowel, cutoff_duration)
# inspect
head(max_durs)
```

overview

```{r}
library(janitor)
tb0 <- fdat %>%
  dplyr::ungroup() %>%
  dplyr::filter(label == "I" | label == "i:" | label == "U" | label == "u:" ) %>%
  dplyr::group_by(type) %>%
  dplyr::mutate(speakers = length(table(speaker))) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(type, label) %>%
  dplyr::summarise(speakers = speakers,
                   obs = n()) %>%
  unique() %>%
  tidyr::spread(label, obs) %>%
  dplyr::ungroup()  %>%
  adorn_totals("row")%>%
  adorn_totals("col") %>%
  dplyr::mutate(Total = Total-speakers)
# save
write.table(tb0, here::here("tables", "tb0.txt"), sep = "\t")
# inspect
tb0
```

remove duration outliers

```{r}
fdat <- dplyr::left_join(fdat, max_durs) %>%
  dplyr::filter(duration <= cutoff_duration,
                ifelse(label != "A:", duration <= 0.4, duration <= 0.75))
# inspect
head(fdat); nrow(fdat)
```




```{r}
dur_IiUu <-  fdat %>%
  dplyr::filter(label == "i:" | label == "I" | label == "u:" | label == "U") %>%
  ggplot(aes(y = vowel, x = duration, fill = type)) +
  geom_density_ridges(alpha = .5) +
  theme_ridges() +
  theme(legend.position = "top") +
  scale_fill_manual(values = c("orange", "gray"))
ggsave(here::here("images", "dur_IiUu.png"), height = 3,  width = 5, dpi = 320)
dur_IiUu
```


```{r}
fdat %>%
  ggplot(aes(x = vowel, y = duration, fill = type)) +
  geom_boxplot()
```



```{r}
fdat %>%
  dplyr::group_by(type, tvariety, vowel) %>%
  dplyr::mutate(ldur = log(duration)) %>%
  ggplot(aes(x = vowel, y = ldur, fill = type)) +
  geom_boxplot() +
  facet_grid(~tvariety)
```


# Remove outliers

```{r}
nrow(fdat)
fdat <- fdat %>% 
  dplyr::group_by(type, tvariety, label) %>%
  dplyr::mutate(densF1 = sm.density(barkF1,eval.points=barkF1)$estimate,
                densF2 = sm.density(barkF2,eval.points=barkF2)$estimate,
                qnt1F1 = quantile(densF1, .5),
                qnt1F2 = quantile(densF2, .5))
# inspect
nrow(fdat); head(fdat); summary(fdat$densF1)
```

```{r}
fdat <- fdat %>%
  # remove data point within the lowest 25 percent of densities among F1 and F2 values
  # but when label = A: and tvariety is gb
  dplyr::mutate(rmv = ifelse(label == "A:" & tvariety == "gb", "keep", "remove")) %>%
  dplyr::rowwise() %>%
  dplyr::filter(rmv == "keep" | densF1 > qnt1F1 | densF2 > qnt1F2) %>%
  # remove superfluous columns
  dplyr::select(-densF1, -densF2, -qnt1F1, -qnt1F2, -cutoff_duration, -rmv)
# inspect
head(fdat); nrow(fdat)
```






```{r sdrmv, eval = F}
fdat <- fdat %>%
  dplyr::filter(lobF1 > -1.96,
                lobF1 < 1.96,
                lobF2 > -1.96,
                lobF2 < 1.96)
```


# Remove GB target

```{r tvar}
fdat <- fdat %>%
  dplyr::filter(tvariety != "gb")
```

# Check density



```{r}
wordplot3 <- function(fdat, vwl1, vwl2){
  plt <- fdat %>% 
    dplyr::filter(label == vwl1 | label == vwl2) %>%
    dplyr::group_by(word, label) %>%
    dplyr::mutate(meanF2 = mean(lobF2),
                  meanF1 = mean(lobF1)) %>%
    dplyr::ungroup() %>%
    dplyr::group_by(label) %>%
    dplyr::mutate(cF2 = mean(lobF2),
                  cF1 = mean(lobF1)) %>%
    # plot
    ggplot(aes(x = lobF2, y = lobF1)) +
  stat_density_2d(geom = "polygon",
                  aes(alpha = ..level.., fill = label), bins = 8)  +
    facet_grid( ~ type) +
    scale_x_reverse(position = "top") + 
    scale_y_reverse(position = "right") + 
    #scale_fill_distiller(palette = "Blues", direction = 1) +
    geom_text(aes(x = meanF2, y = meanF1, 
                  label = word, color = label), size = 4) +
    geom_text(aes(x = cF2, y = cF1, 
                  label = label), size = 6, color = "gray20") +
    theme_minimal() +
    theme(panel.grid.major = element_blank(), 
                  panel.grid.minor = element_blank(),
                  legend.position = "none") +
  scale_color_manual(values = c("orange3", "gray40")) +
  scale_fill_manual(values = c("orange", "gray")) +
    labs(x = "Formant 2 (Lobanov normalized)", y = "Formant 1 (Lobanov normalized)")
  return(plt)
  }
```


```{r}
pIi <- wordplot3(fdat = fdat, vwl1 = "I", vwl2 = "i:")
ggsave(here::here("images", "pIi.png"), height = 3,  width = 5, dpi = 320)
pIi
```

```{r}
pUu <- wordplot3(fdat = fdat, vwl1 = "U", vwl2 = "u:")
ggsave(here::here("images", "pUu.png"), height = 3,  width = 5, dpi = 320)
pUu
```





# Normalization

See [here](https://marissabarlaz.github.io/portfolio/vowelnormalization/)

Gender normalization

```{r}
# Using mean F3 of open vowels (vowels with F1>600) to approximate female F1&F2 to those of male
tempvowel <- fdat %>% 
  dplyr::filter(F1 > 600) %>% 
  dplyr::group_by(gender) %>%
  dplyr::mutate(meanF3 = mean(F3))
# extract k: k = ratio of Female F3 to Male F3
k <- max(tempvowel$meanF3) / min(tempvowel$meanF3)
# perform normalization
fdat <- fdat %>%
  dplyr::mutate(barkF1 = phonR::normBark(F1),
                barkF2 = phonR::normBark(F2)) %>%
  dplyr::mutate(normF1 = dplyr::if_else(gender == "f", F1*k, F1),
                normF2 = dplyr::if_else(gender == "f", F2*k, F2))
```

calculate distance from target vowel centroid for each token

```{r}
tcentroid <- fdat %>%
  dplyr::filter(type == "ENS") %>%
  dplyr::group_by(tvariety, vowel) %>%
  dplyr::summarise(cF1 = median(barkF1),
                   cF2 = median(barkF2))
# inspect
head(tcentroid)
```

combine

```{r}
fdat <- dplyr::left_join(fdat, tcentroid) %>%
  # euclidean distance
  dplyr::rowwise() %>%
  dplyr::mutate(ED = sqrt((barkF1-cF1)^2 + (barkF2-cF2)^2)) %>%
  # remove superfluous columns
  dplyr::select(-remove)
# inspect
head(fdat)
```

# Gramm. Status of Word

```{r}
gram <- c("be", "but", "did", "do", "get", "gets", "got", "had", "has", "he", "his", 
          "she", "that", "this", "too", "was", "we", "what", "who", "with")
fdat <- fdat %>%
  # gramm. category of word
  dplyr::mutate(WordType = ifelse(word %in% gram, "grammatical", "lexical"))
```




```{r, eval = T}
fdat %>%
  dplyr::group_by(label) %>%
  dplyr::mutate(meanF2 = median(barkF2),
                meanF1 = median(barkF1)) %>%
  ggplot(aes(x = barkF2, y = barkF1, color = vowel, fill = vowel, label = vowel)) +
  facet_grid(tvariety ~ type) +
  scale_x_reverse(position = "top") + 
  scale_y_reverse(position = "right") + 
  #stat_ellipse(aes(group = label), level = .5) +
  #geom_point(alpha = .2) +
  geom_text(alpha = .2) +
  #geom_count() +
  #geom_text(aes(x = meanF2, y = meanF1, color = vowel, label = vowel), size = 5) +
  #scale_color_viridis(discrete = TRUE) +
  theme_minimal() +
  theme(legend.position = "right") +
  labs(x = "Formant 2 (Bark normalized)", y = "Formant 1 (Bark normalized))")
ggsave(file = here("images", "vowels_lob.png"), height = 5,  width = 5, dpi = 320)
```



# Overview of the data

```{r}
tb1 <- fdat %>%
  dplyr::ungroup() %>%
  dplyr::filter(label == "I" | label == "i:" | label == "U" | label == "u:" ) %>%
  dplyr::group_by(type) %>%
  dplyr::mutate(speakers = length(table(speaker))) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(type, label) %>%
  dplyr::summarise(speakers = speakers,
                   obs = n()) %>%
  unique() %>%
  tidyr::spread(label, obs) %>%
  dplyr::ungroup()  %>%
  adorn_totals("row")%>%
  adorn_totals("col") %>%
  dplyr::mutate(Total = Total-speakers)
# save
write.table(tb1, here::here("tables", "tb1.txt"), sep = "\t")
# inspect
tb1
```

```{r}

tb2 <- fdat %>%
  dplyr::ungroup() %>%
  dplyr::filter(label == "I" | label == "i:" | label == "U" | label == "u:" ) %>%
  dplyr::mutate(Age = dplyr::case_when(Age < 30 ~ "18-29",
                                       Age < 40 ~ "30-39",
                                       Age < 50 ~ "40-49",
                                       Age > 49 ~ "50+",
                                       TRUE ~ "unknown")) %>%
  dplyr::group_by(type, gender, Age) %>%
  dplyr::summarise(speakers = length(table(speaker))) %>%
  tidyr::spread(Age, speakers) %>%
  dplyr::ungroup()  %>%
  adorn_totals("row")%>%
  adorn_totals("col")
# save
write.table(tb2, here::here("tables", "tb2.txt"), sep = "\t")
# inspect
tb2
```

tabulate proficiency


```{r}
tb3 <- fdat %>%
  dplyr::ungroup() %>%
  dplyr::filter(label == "I" | label == "i:" | label == "U" | label == "u:" ) %>%
  dplyr::filter(type == "JPN")%>%
  dplyr::group_by(prof, gender) %>%
  dplyr::summarise(speakers = length(table(speaker))) %>%
  tidyr::spread(prof, speakers) %>%
  dplyr::ungroup()  %>%
  adorn_totals("row")%>%
  adorn_totals("col")
# save
write.table(tb3, here::here("tables", "tb3.txt"), sep = "\t")
# inspect
tb3
```


# Save data to disc

```{r savefdat, message=F, warning=F}
# save
base::saveRDS(fdat, file = here::here("data", "fdat_sst.rda"))
# inspect
nrow(fdat); head(fdat)
```


# Statistical Analysis

## Preparation

install packages

```{r install, eval=F, message=F, warning=F}
# install
install.packages("tidyverse")
install.packages("here")
install.packages("adehabitatHR")
install.packages("lme4")
install.packages("sjPlot")
install.packages("report")
install.packages("flextable")
```

load packages

```{r load, message=F, warning=F}
library(tidyverse)
library(here)
library(adehabitatHR)
library(lme4)
library(sjPlot)
library(report)
library(flextable)
```


load data


```{r data, message=F, warning=F}
# load .rda data
fdat  <- base::readRDS(file = here::here("data", "fdat_sst.rda"))
# inspect
nrow(fdat); head(fdat)
```

remove superfluous variables

```{r}
fdat <- fdat %>%
  dplyr::select(id, label, word, file, F1, F2, speaker, vowel, 
                label, type, gender, Age, tvariety, edist, barkF1, 
                barkF2, fspeaker, lobF1, lobF2, duration, normF1, 
                normF2, cF1, cF2, ED, WordType, prof) %>%
  # create new speaker column
  dplyr::mutate(speaker = paste0(type, speaker))
# inspect
head(fdat)
```



# Bhattacharyya's affinity 

function for extracting Bhattacharyya's affinity by type and target variety

```{r bafun, message=F, warning=F}
exba <- function(data, section, target, vwl1, vwl2){
  ba <- data %>%
  ungroup() %>%
  dplyr::filter(type == section,
                tvariety == target,
                label == vwl1 | label == vwl2)

  ba_formants <- ba %>%  dplyr::select(lobF1, lobF2)
  # extract vowels
  ba_vowel <- ba %>%  dplyr::select(vowel)
  # spatial data frame
  ba_spdf <- SpatialPointsDataFrame(ba_formants, ba_vowel)
  # calculate Bhattacharyya's affinity
  ba_ba <- kerneloverlap(ba_spdf, method = "BA")
  # result
  return(ba_ba[1,2])
}
```



## I vs i: 


```{r baius, message=F, warning=F}
# JPN
exba(data = fdat, section = "JPN", target = "us", vwl1 = "I", vwl2 = "i:")
# ENS
exba(data = fdat, section = "ENS", target = "us", vwl1 = "I", vwl2 = "i:")
```



## U vs u:

High-back 

```{r baius, message=F, warning=F}
# JPN
exba(data = fdat, section = "JPN", target = "us", vwl1 = "U", vwl2 = "u:")
# ENS
exba(data = fdat, section = "ENS", target = "us", vwl1 = "U", vwl2 = "u:")
```




# Duration

```{r durtb, message=F, warning=F}
durtb <- fdat %>%
  dplyr::group_by(type, tvariety, vowel) %>%
  dplyr::summarise(duration_mean = round(mean(duration), 3), 
                   duration_sd = round(sd(duration), 3),
                   freq = n())
# inspect
durtb
```


```{r}
durtb %>%
  dplyr::mutate(mean_sd = paste0(duration_mean, " (", duration_sd, ", N = ", freq, ")")) %>%
  dplyr::select(-duration_mean, -duration_sd, -freq) %>%
  tidyr::spread(type, mean_sd) %>%
  flextable::flextable()
```


```{r durplot, message=F, warning=F}
fdat %>%
  dplyr::group_by(type, tvariety, vowel) %>%
  dplyr::summarise(duration = mean(duration)) %>%
  ggplot(aes(x = vowel, y = duration, fill = type)) +
  geom_bar(stat = "identity", position=position_dodge()) +
  facet_grid( ~ tvariety) +
  theme_bw()
```



```{r}
diff_dat <- fdat %>%
  dplyr::group_by(type, tvariety, label) %>%
  dplyr::summarise(dur = mean(duration)) %>%
  dplyr::filter(label == "I" | 
                  label == "i:" | 
                  label == "U" | 
                  label == "u:") %>%
  tidyr::spread(label, dur) %>%
  dplyr::mutate(diff_i = `i:` - I,
                diff_u = `u:` - U) %>%
  dplyr::select(-I, -U, -`i:`, -`u:`) %>%
  tidyr::gather(pair, diff, diff_i:diff_u) %>%
  dplyr::mutate(pair = dplyr::case_when(pair == "diff_i" ~ "Ii",
                                        pair == "diff_u" ~ "Uu"))
# inspect
diff_dat
```


```{r eval = F}
diff_dat <- fdat %>%
  dplyr::group_by(type, tvariety, label) %>%
  dplyr::summarise(dur = round(mean(duration), 5),
                   sd = round(sd(duration), 5)) %>%
  dplyr::filter(label == "I" |
                  label == "i:" | 
                  label == "U" | 
                  label == "u:") %>%
  tidyr::spread(label, dur) %>%
  dplyr::mutate(diff_i = `i:` - I,
                diff_u = `u:` - U) %>%
  dplyr::select(-I, -U, -`i:`, -`u:`) %>%
  tidyr::gather(pair, diff, diff_i:diff_u) %>%
  dplyr::mutate(pair = dplyr::case_when(pair == "diff_i" ~ "Ii",
                                        pair == "diff_u" ~ "Uu"))
# inspect
diff_dat
```




```{r}
dplot <- diff_dat %>%
  dplyr::rename(Speaker = type) %>%
  dplyr::mutate(pair = ifelse(pair == "Ii", "i-vowles", "u-vowels")) %>%
  ggplot(aes(x = pair, y = diff, fill = Speaker)) +
  geom_bar(stat = "identity", position = position_dodge()) +
    # define colors
  scale_fill_manual(values= c("orange", "gray")) + 
  # add text and define colour
  geom_text(aes(label=round(diff, 3)), vjust=-1.6, color="gray20", 
            # define text position and size
            position = position_dodge(0.9),  size=3) +
  theme_bw() +
  coord_cartesian(ylim = c(0, .06)) +
  labs(x = "Vowel pairs", y = "Mean Difference in Duration (ms)") +
  theme(legend.position = "top")
# save
ggsave(here::here("images", "dplot.png"), height = 3,  width = 5, dpi = 320)
# inspect
dplot
```


# Statistical Analysis

base-line model

```{r}
m0 <- lmer(duration ~ 1 + (1 | speaker) + (1 | word), data = fdat)
# inspect
summary(m0)
```

## Model Fitting


**Main effects**

### type

sig.

```{r}
# add type
m1 <- update(m0, . ~ . + type)
# inspect
sjPlot::tab_model(m1)
```

### label (vowel)

sig.

```{r}
# add label
m2 <- update(m1, . ~ . + label)
# checks
anova(m1, m2, test = "Chi")
car::vif(m2)
# inspect
sjPlot::tab_model(m2)
```

### gender

not sig.

```{r}
# add gender
m3 <- update(m2, . ~ . + gender)
# checks
anova(m2, m3, test = "Chi")
car::vif(m3)
# inspect
sjPlot::tab_model(m3)
```

### Age

not sig.


```{r}
# add Age
m4 <- update(m2, . ~ . + Age)
# checks
anova(m2, m4, test = "Chi")
car::vif(m4)
# inspect
sjPlot::tab_model(m4)
```

### WordType

not sig.

```{r}
# add WordType
m5 <- update(m2, . ~ . + WordType)
# checks
anova(m2, m5, test = "Chi")
car::vif(m5)
# inspect
sjPlot::tab_model(m5)
```



**Interactions**

### type x label

sig.

```{r}
# add type x label
m6 <- update(m2, . ~ . + type * label)
# checks
anova(m2, m6, test = "Chi")
car::vif(m6)
# inspect
sjPlot::tab_model(m6)
```


### type x Age

unacc. vifs

```{r}
# add type x Age
m7 <- update(m6, . ~ . + type * Age)
# checks
car::vif(m7)
```

### type x gender

unacc. vifs

```{r}
# add type x gender
m8 <- update(m6, . ~ . + type * gender)
# checks
car::vif(m8)
```


### type x WordType

unacc. vifs

```{r}
# add type x gender
m9 <- update(m6, . ~ . + type * WordType)
# checks
car::vif(m9)
```

### label x gender

sig.

```{r}
# add type x gender
m10 <- update(m6, . ~ . + type * gender)
# checks
car::vif(m10)
anova(m6, m10, test = "Chi")
# inspect
sjPlot::tab_model(m10)
```


### label x Age

unacc. vifs

```{r}
# add type x gender
m11 <- update(m10, . ~ . + type * Age)
# checks
car::vif(m11)
```



### label x WordType

unacc. vifs

```{r}
# add type x gender
m12 <- update(m10, . ~ . + type * WordType)
# checks
car::vif(m12)
```

### gender x Age

unacc. vifs

```{r}
# add type x gender
m13 <- update(m10, . ~ . + gender * Age)
# checks
car::vif(m13)
```

### gender x WordType

not sig

```{r}
# add type x gender
m14 <- update(m10, . ~ . + gender * WordType)
# checks
car::vif(m14)
anova(m10, m14, test = "Chi")
# inspect
sjPlot::tab_model(m14)
```

### Age x WordType

not sig.

```{r}
# add type x gender
m15 <- update(m10, . ~ . + Age * WordType)
# checks
car::vif(m15)
anova(m10, m15, test = "Chi")
# inspect
sjPlot::tab_model(m15)
```

## Final Minimal Adequate Model

```{r}
sjPlot::tab_model(m0, m10)
```


```{r}
sjPlot::tab_model(m10)
```

```{r}
sjPlot::plot_model(m10) +
  coord_flip(ylim = c(-.1, .1)) 
```


```{r}
plot_model(m10, type = "re", terms = "words")
```


```{r}
fixef(m10)
```


```{r}
# jpn i
jpn_I <- fixef(m10)[1] + fixef(m10)[2]
jpn_i <- fixef(m10)[1] + fixef(m10)[2] + fixef(m10)[3]  + fixef(m10)[7]
# ens i
ens_I <- fixef(m10)[1] 
ens_i <- fixef(m10)[1] + fixef(m10)[3]

# jpn u
jpn_U <- fixef(m10)[1] + fixef(m10)[2] + fixef(m10)[4]  + fixef(m10)[8]
jpn_u <- fixef(m10)[1] + fixef(m10)[2] + fixef(m10)[5]  + fixef(m10)[9]
# ens i
ens_U <- fixef(m10)[1] + fixef(m10)[4]
ens_u <- fixef(m10)[1] + fixef(m10)[5]
```




```{r efp, message=F, warning=F}
plot_model(m10, type = "pred", terms = c("type", "label"), axis.lim = c(0, .2)) +
  theme_bw() +
  labs(x = "Speaker type", y = "Predicted Duration (ms)") +
  geom_segment(aes(x = 1.9, y = jpn_I, xend = 1.9, yend = jpn_i), 
               size = 1, color = "gray") +
  geom_segment(aes(x = 0.9, y = ens_I, xend = 0.9, yend = ens_i), 
               size = 1, color = "gray") +
    geom_segment(aes(x = 2.1, y = jpn_U, xend = 2.1, yend = jpn_u), 
                 size = 1, color = "orange") +
  geom_segment(aes(x = 1.1, y = ens_U, xend = 1.1, yend = ens_u), 
               size = 1, color = "orange") +
  scale_color_manual(values = c("lightgray", "darkgray", "orange", "orange3"))
# save
ggsave(here::here("images", "ef.png"), height = 3,  width = 5, dpi = 320)
```






```{r}
report::report(m10)
```



# Citation & Session Info

Schweinberger, Martin. `r format(Sys.time(), '%Y')`. A Corpus-Based Acoustic Analysis of Monophthongal Vowels among Japanese Learners and Native Speakers of English. Brisbane: The University of Queensland, School of Languages and Cultures. url: https://slcladal.github.io/praatrf.html (Version `r format(Sys.time(), '%Y.%m.%d')`).

@manual{schweinberger`r format(Sys.time(), '%Y')`praatrf,
  author = {Schweinberger, Martin},
  title = {A Corpus-Based Acoustic Analysis of Monophthongal Vowels among Japanese Learners and Native Speakers of English},
  note = {https://slcladal.github.io/praatrf.html},
  year = {`r format(Sys.time(), '%Y')`},
  organization = "The University of Queensland, School of Languages and Cultures},
  address = {Brisbane},
  edition = {`r format(Sys.time(), '%Y.%m.%d')`}
}


```{r}
sessionInfo()
```
