---
title: "Reproducibility in corpus-based computational analyses of learner speech - Part 1"
author: "Martin Schweinberger"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: word_document
---

# Introduction

This R Notebook shows how to access Praat from R and how to extract formant values from wav files and their corresponding Praat TextGrids. The tutorials is based on [Phonetics Tools in R](https://marissabarlaz.github.io/portfolio/phoneticstools/] by Marissa Barlaz).

## Preparation

Prepare session

```{r prep0}
# set options
options(stringsAsFactors = F)         # no automatic data transformation
options("scipen" = 100, "digits" = 4) # suppress math annotation
```

Install packages

```{r install,eval=F}
install.packages("here")
install.packages("tidyverse")
install.packages("rPraat")
install.packages("phonTools")
install.packages("tuneR")
install.packages("ipa")
install.packages("wrassp")
# `PraatR` is not available on CRAN at the time of writing 
# thus it has to be installed from GitHub (as shown below)
install.packages("devtools")
devtools::install_github("usagi5886/PraatR")
```


Load packages

```{r prep1, warning = F, message = F}
library(here)
library(tidyverse)
library(rPraat)
library(phonTools)
library(tuneR)
library(ipa)
library(wrassp)
library(PraatR)
```



# Formant Extraction

# Function for extraction

```{r extract, message = F, warning = F}
# apply function
exformants <- function(paths, formants, gender, tvar){
  sapply(paths, function(x){
     # define pattern
    target <- c("\\b[gkbpdtwhs]{1,}[aeiouyw]{1,}'{0,1}[gkdtbpkhs]{0,3}\\b")
    # extract formants from file
    PathWav <- x
    PathStore = stringr::str_replace_all(PathWav, ".txt", paste0("_F", formants, "_", gender, ".rda"))
    PathStore = stringr::str_replace_all(PathStore, "/data/[JPNENS]{3,3}_{0,1}[A-Z]{0,2}/", "/tables/")
    PathTextGrid = stringr::str_replace(PathWav, ".txt", ".TextGrid")
    PathWav = stringr::str_replace(PathWav, ".txt", ".wav")
    TextGridInfo = tg.read(PathTextGrid)
    T1 = data.frame(tmin = TextGridInfo[1][[1]]$t1,
                    word = TextGridInfo[1][[1]]$label)
    T3 = data.frame(tmin =TextGridInfo$MAU$t1,
                    tmax = TextGridInfo$MAU$t2,
                    label = TextGridInfo$MAU$label) %>% 
      dplyr::filter(label !="")
    TextGrid = dplyr::left_join(T3, T1, "tmin") %>%
      tidyr::fill(word)  %>% 
      # simplify data frame
      dplyr::mutate(file = stringr::str_replace_all(PathWav, ".*/(.*?).wav", "\\1")) %>%
      dplyr::filter(stringr::str_detect(word, target)) %>%
      dplyr::filter(stringr::str_detect(label, "i|I|e|E|\\{|y|2|9|1|\\@|6|3|a|\\}|8|\\&|M|7|V|A|u|U|o|O|Q")) %>%
      dplyr::mutate(midpoint = (tmin+tmax)/2) %>%
      # extract formants
      dplyr::rowwise() %>%
      dplyr::mutate(F1 = mean(wrassp::forest(PathWav, toFile = F, beginTime = tmin,
                                         endTime = tmax, numFormants = formants,
                                         gender = gender)$fm[,1]),
                    F2 = mean(wrassp::forest(PathWav, toFile = F,beginTime = tmin, 
                                         endTime = tmax, numFormants = formants,
                                         gender = gender)$fm[,2]),
                    F3 = mean(wrassp::forest(PathWav, toFile = F, beginTime = tmin,
                                         endTime = tmax, numFormants = formants,
                                         gender = gender)$fm[,3])) %>%
      dplyr::mutate(vowel = sampa(label, to = "ipa")) %>%
      dplyr::mutate(path = paste0(str_replace_all(PathStore, ".rda", "_"), tvar, collapse = ""))
    PathStore = paste0(str_replace_all(PathStore, ".rda", "_"), tvar, ".rda", collapse = "")
    # save textgrid
    base::saveRDS(TextGrid, file = PathStore)
  })
}
```


# Extract JPN vowels

Find files with target words

```{r texts, message=F, warning=F}
txtfiles <- list.files(here("data/JPN_TU"), ".*.txt", full.names = T) 
txts <- sapply(txtfiles, function(x) {
  x <- scan(x, what = "char", quiet = T) %>%
    paste0(collapse = " ") %>%
  str_squish()
}) 
# define pattern
target <- c("\\b[gkbpdtwhs]{1,}[aeiouyw]{1,}'{0,1}[gkdtbpkhs]{0,3}\\b")
targetfiles <- txtfiles[stringr::str_detect(txts, target)]
# inspect
head(targetfiles, 10)

```


## Target US

```{r jmus, eval = T, message = F, warning=F}
## JPN Male US
# gender = male, formants = 3, target variety = genam
exformants(paths = targetfiles, formants = 3, gender = "m", tvar = "us")
# gender = male, formants = 4, target variety = genam
exformants(paths = targetfiles, formants = 4, gender = "m", tvar = "us")
# gender = male, formants = 5, target variety = genam
exformants(paths = targetfiles, formants = 5, gender = "m", tvar = "us")
# gender = male, formants = 6, target variety = genam
exformants(paths = targetfiles, formants = 6, gender = "m", tvar = "us")
# gender = male, formants = 7, target variety = genam
exformants(paths = targetfiles, formants = 7, gender = "m", tvar = "us")

## JPN Female US
# gender = female, formants = 3, target variety = genam
exformants(paths = targetfiles, formants = 3, gender = "f", tvar = "us")
# gender = female, formants = 4, target variety = genam
exformants(paths = targetfiles, formants = 4, gender = "f", tvar = "us")
# gender = female, formants = 5, target variety = genam
exformants(paths = targetfiles, formants = 5, gender = "f", tvar = "us")
# gender = female, formants = 6, target variety = genam
exformants(paths = targetfiles, formants = 6, gender = "f", tvar = "us")
# gender = female, formants = 7, target variety = genam
exformants(paths = targetfiles, formants = 7, gender = "f", tvar = "us")
```


## Target GB

Find files with target words

```{r texts, message=F, warning=F}
txtfiles <- list.files(here("data/JPN_TB"), ".*.txt", full.names = T) 
txts <- sapply(txtfiles, function(x) {
  x <- scan(x, what = "char", quiet = T) %>%
    paste0(collapse = " ") %>%
  str_squish()
}) 
# define pattern
target <- c("\\b[gkbpdtwhs]{1,}[aeiouyw]{1,}'{0,1}[gkdtbpkhs]{0,3}\\b")
targetfiles <- txtfiles[stringr::str_detect(txts, target)]
# inspect
head(targetfiles, 10)

```

```{r jmgb, eval = T, message = F, warning=F}
## JPN Male GB
# gender = male, formants = 3, target variety = gb
exformants(paths = targetfiles, formants = 3, gender = "m", tvar = "gb")
# gender = male, formants = 4, target variety = gb
exformants(paths = targetfiles, formants = 4, gender = "m", tvar = "gb")
# gender = male, formants = 5, target variety = gb
exformants(paths = targetfiles, formants = 5, gender = "m", tvar = "gb")
# gender = male, formants = 6, target variety = gb
exformants(paths = targetfiles, formants = 6, gender = "m", tvar = "gb")
# gender = male, formants = 7, target variety = gb
exformants(paths = targetfiles, formants = 7, gender = "m", tvar = "gb")

## JPN Female GB
# gender = female, formants = 3, target variety = gb
exformants(paths = targetfiles, formants = 3, gender = "f", tvar = "gb")
# gender = female, formants = 4, target variety = gb
exformants(paths = targetfiles, formants = 4, gender = "f", tvar = "gb")
# gender = female, formants = 5, target variety = gb
exformants(paths = targetfiles, formants = 5, gender = "f", tvar = "gb")
# gender = female, formants = 6, target variety = gb
exformants(paths = targetfiles, formants = 6, gender = "f", tvar = "gb")
# gender = female, formants = 7, target variety = gb
exformants(paths = targetfiles, formants = 7, gender = "f", tvar = "gb")
```

# Extract ENS vowles

Find files with target words

```{r texts, message=F, warning=F}
txtfiles <- list.files(here("data/ENS"), ".*.txt", full.names = T) 
txts <- sapply(txtfiles, function(x) {
  x <- scan(x, what = "char", quiet = T) %>%
    paste0(collapse = " ") %>%
  str_squish()
}) 
# define pattern
target <- c("\\b[gkbpdtwhs]{1,}[aeiouyw]{1,}'{0,1}[gkdtbpkhs]{0,3}\\b")
targetfiles <- txtfiles[stringr::str_detect(txts, target)]
# inspect
head(targetfiles, 10)

```

```{r eve, eval = T, message = F, warning=F}
## ENS Male
# gender = male, formants = 3
exformants(paths = targetfiles, formants = 3, gender = "m", tvar = "ve")
# gender = male, formants = 4
exformants(paths = targetfiles, formants = 4, gender = "m", tvar = "ve")
# gender = male, formants = 5
exformants(paths = targetfiles, formants = 5, gender = "m", tvar = "ve")
# gender = male, formants = 6
exformants(paths = targetfiles, formants = 6, gender = "m", tvar = "ve")
# gender = male, formants = 7
exformants(paths = targetfiles, formants = 7, gender = "m", tvar = "ve")

## ENS Female
# gender = male, formants = 3
exformants(paths = targetfiles, formants = 3, gender = "f", tvar = "ve")
# gender = male, formants = 4
exformants(paths = targetfiles, formants = 4, gender = "f", tvar = "ve")
# gender = male, formants = 5
exformants(paths = targetfiles, formants = 5, gender = "f", tvar = "ve")
# gender = male, formants = 6
exformants(paths = targetfiles, formants = 6, gender = "f", tvar = "ve")
# gender = male, formants = 7
exformants(paths = targetfiles, formants = 7, gender = "f", tvar = "ve")
```


# Combine

```{r combine, message=F, warning=F}
tbpaths <- list.files(here::here("tables"), pattern = ".rda", full.names = T)
ltb <- lapply(tbpaths, function(x){
  base::readRDS(x)
})
vdf <- bind_rows(ltb, .id = "id")
# save
base::saveRDS(vdf, file = here::here("data", "vdf.rda"))
# inspect
nrow(vdf); head(vdf)
```



# Citation & Session Info

Schweinberger, Martin. `r format(Sys.time(), '%Y')`. Reproducibility in corpus-based computational analyses of learner speech. Brisbane: The University of Queensland, School of Languages and Cultures. url: https://slcladal.github.io/praatrf.html (Version `r format(Sys.time(), '%Y.%m.%d')`).

@manual{schweinberger`r format(Sys.time(), '%Y')`praatrf,
  author = {Schweinberger, Martin},
  title = {Reproducibility in corpus-based computational analyses of learner speech},
  note = {https://slcladal.github.io/praatrf.html},
  year = {`r format(Sys.time(), '%Y')`},
  organization = "The University of Queensland, School of Languages and Cultures},
  address = {Brisbane},
  edition = {`r format(Sys.time(), '%Y.%m.%d')`}
}


```{r}
sessionInfo()
```
